### 2021-07-21 paper/localpsf.tex
edits to rational positive definite section based on feedback from J

In many problems of practical interest (e.g., Hessian approximation) $\mathcal{A}$ is symmetric positive semi-definite. However, $\widetilde{\mathbf{A}}$ is generally non-symmetric and indefinite because of errors in the product-convolution approximation. 
This is undesirable. Symmetry and positive semi-definiteness are important properties which should be preserved by the approximation, if possible. Also, a lack of symmetry and/or positive semi-definiteness in $\widetilde{\mathbf{A}}$ may prevent one from using popular and highly effective algorithms, such as the conjugate gradient method, to perform further useful operations involving $\widetilde{\mathbf{A}}$.
In this section we describe a method for modifying $\widetilde{\mathbf{A}}$ to make it symmetric positive semi-definite.

First, we use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 \mathbf{I} + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the spectrum of $\widetilde{\mathbf{A}}_\text{sym}^+$ has certain desirable properties. Later in this section we will describe how $c_0$, $c_1$, $c_2$, and $\mu$ are chosen and how our choice of these scalars affects the spectral properties of $\widetilde{\mathbf{A}}_\text{sym}^+$. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing an H-matrix inverse, $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is fine to compute the inverse because $\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}$ is an H-matrix. There exist fast, stable, scalable, and widely used algorithms for inverting H-matrices \cite{Hmatrixinverse}.

The rational matrix approximation \eqref{eq:rational_matrix_function} may be written as $\widetilde{\mathbf{A}}_\text{sym}^+ = f(\widetilde{\mathbf{A}}_\text{sym})$, where $f$ is the rational function 
\begin{equation}
\label{eq:rational_function_scalar}
f(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
We choose the four unknown constants, $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of four equations:
\begin{align}
f(0) &= 0 \label{eq:desired_property_1} \\
f'(0) &= 0 \label{eq:desired_property_2} \\
f(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
f(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
\end{align}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, respectively. We also require $\mu > |\lambda_\text{min}|$, so that the pole of $f$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. Note that $\lambda_\text{min} < 0$ since $\widetilde{\mathbf{A}}_\text{sym}$ is indefinite. 
We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}, a matrix-free Krylov method. This is fast because computing matrix-vector products with an H-matrix is fast, and because the eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ do not cluster near $\lambda_\text{min}$ or $\lambda_\text{max}$ (though the eigenvalues may cluster at zero).

@@

Using the definition of $f$ in \eqref{eq:rational_function_scalar}, and basic calculus and linear algebra, we may write Equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, and \eqref{eq:desired_property_4} in the following matrix form:
\begin{equation}
\label{eq:rational_3_linear_system}
\begin{bmatrix} 1 & 0 & 1/\mu \\ 0 & 1 & -1/\mu^2 \\ 1 & \lambda_\text{max} & 1/(\lambda_\text{max}+\mu) \end{bmatrix} \begin{bmatrix} c_0 \\ c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \\ \lambda_\text{max} \end{bmatrix},
\end{equation}
and we may write Equation \eqref{eq:desired_property_3} as follows: 
\begin{equation}
\label{eq:rational_extra_eq}
c_0 + c_1 \lambda_\text{min} + \frac{c_2}{\lambda_\text{min}+\mu} - |\lambda_\text{min}| = 0.
\end{equation}
Let $g(\mu)$ denote the left hand side of \eqref{eq:rational_extra_eq}, but with $c_0$, $c_1$, and $c_2$ now solving \eqref{eq:rational_3_linear_system} for the given value of $\mu$. I.e., within $g(\mu)$ the quantities $c_0$, $c_1$, and $c_2$ are no longer free variables; they now depend on $\mu$ implicitly. We solve the combined system, \eqref{eq:rational_3_linear_system} and $\eqref{eq:rational_extra_eq}$, by solving the one-dimensional nonlinear root finding problem
\begin{equation}
\label{eq:g_mu_zero}
g(\mu)=0
\end{equation}
with bound $\mu > |\lambda_\text{min}|$ and initial guess $\mu_0 = 2 |\lambda_\text{min}|$. Each evaluation of $g$ requires solving the $3 \times 3$ linear system \eqref{eq:rational_3_linear_system}, and once the solution $\mu$ is found, we solve \eqref{eq:rational_3_linear_system} one more time to get the final values of $c_0$, $c_1$, and $c_2$.
%
%Row 1 corresponds to the condition $f(M)=M$, row 2 corresponds to the condition $f(0)=0$, and row 3 corresponds to the condition $f'(0)=0$. To find the third row of $A$, we used the fact that the derivative of $f$ is given by $$f'(\lambda) = c_1 - \frac{c_2}{(\lambda+\mu)^2}.$$
%
%
%We want $f(\lambda)$ to have the following desirable properties:
%\begin{enumerate}
%	\item $f(\lambda)$ is nonnegative on $[m, M]$.
%	\item $f(\lambda) \approx \lambda$ when $\lambda$ is positive and not small.
%	\item $f(0)=0$ is the minimum of $f(\lambda)$ on $[m, M]$.
%	\item $f(\lambda) \le |\lambda|$ when $\lambda$ is negative.
%\end{enumerate}
%Property 1 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Property 2 ensures that $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. That is, the part of the spectrum of $\mathcal{A}$ that we expect to be well-approximated by the product-convolution approximation. Property 3 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ preserves the null-space (and near null-space) of $\widetilde{\mathbf{A}}_\text{sym}$. This is important because in practice, we often want to approximate operators $\mathcal{A}$ that have a large (possibly infinite) number of zero eigenvalues, and/or a sequence of eigenvalues that clusters at zero. Hessians in ill-posed inverse problems typically have both of these properties. Property 4 ensures that $f$ does not amplify erroneous negative eigenvalues $\widetilde{\mathbf{A}}_\text{sym}$. If $\mathcal{A}$ is positive semi-definite, then eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ can only be negative because of errors in the product-convolution approximation.

>>>>

In many problems of practical interest (e.g., Hessian approximation) $\mathcal{A}$ is symmetric positive semi-definite. However, $\widetilde{\mathbf{A}}$ is generally non-symmetric and indefinite because of errors in the product-convolution approximation. 
This is undesirable. Symmetry and positive semi-definiteness are important properties which should be preserved if possible. Also, lacking these properties may prevent one from using popular and highly effective algorithms, such as the conjugate gradient method, to perform further useful operations involving $\widetilde{\mathbf{A}}$.
In this section we describe a method for modifying $\widetilde{\mathbf{A}}$ to make it symmetric positive semi-definite.

First, we use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 \mathbf{I} + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ approximately equal the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, while the negative and small positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ are modified so that $\widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Later in this section we will explain how $c_0$, $c_1$, $c_2$, and $\mu$ are chosen and how our choice of these scalars affects the spectral properties of $\widetilde{\mathbf{A}}_\text{sym}^+$. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing an H-matrix inverse, $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is fine to compute the inverse because $\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}$ is an H-matrix. There exist fast, stable, scalable, and widely used algorithms for inverting H-matrices \cite{Hmatrixinverse}.

Approximation \eqref{eq:rational_matrix_function} may be written as $\widetilde{\mathbf{A}}_\text{sym}^+ = f(\widetilde{\mathbf{A}}_\text{sym})$, where
\begin{equation}
\label{eq:rational_function_scalar}
f(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
We choose $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of equations:
\begin{align}
f(0) &= 0 \label{eq:desired_property_1} \\
f'(0) &= 0 \label{eq:desired_property_2} \\
f(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
f(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
\end{align}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, respectively. We also require $\mu > |\lambda_\text{min}|$, so that the pole of $f$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. Note that $\lambda_\text{min} < 0$ since $\widetilde{\mathbf{A}}_\text{sym}$ is indefinite. 
We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}, a matrix-free Krylov method. This is fast because computing matrix-vector products with an H-matrix is fast, and because the eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ do not cluster near $\lambda_\text{min}$ or $\lambda_\text{max}$ (though the eigenvalues may cluster at zero).

@@

Using the definition of $f$ in \eqref{eq:rational_function_scalar}, and basic calculus and linear algebra, we may write Equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, and \eqref{eq:desired_property_4} in the following matrix form:
\begin{equation}
\label{eq:rational_3_linear_system}
\begin{bmatrix} 1 & 0 & 1/\mu \\ 0 & 1 & -1/\mu^2 \\ 1 & \lambda_\text{max} & 1/(\lambda_\text{max}+\mu) \end{bmatrix} \begin{bmatrix} c_0 \\ c_1 \\ c_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \\ \lambda_\text{max} \end{bmatrix},
\end{equation}
and we may write Equation \eqref{eq:desired_property_3} as follows: 
\begin{equation}
\label{eq:rational_extra_eq}
\underbrace{c_0 + c_1 \lambda_\text{min} + \frac{c_2}{\lambda_\text{min}+\mu} - |\lambda_\text{min}|}_{g(\mu)} = 0.
\end{equation}
Let $g(\mu)$ denote the left hand side of \eqref{eq:rational_extra_eq}, but with $c_0$, $c_1$, and $c_2$ being the solution of \eqref{eq:rational_3_linear_system} for the given value of $\mu$. I.e., within $g(\mu)$ the quantities $c_0$, $c_1$, and $c_2$ are no longer free variables; they now depend on $\mu$ implicitly through the solution of a linear system. We solve the combined system, \eqref{eq:rational_3_linear_system} and $\eqref{eq:rational_extra_eq}$, by solving the one-dimensional nonlinear root finding problem
\begin{equation}
\label{eq:g_mu_zero}
g(\mu)=0
\end{equation}
with bound $\mu > |\lambda_\text{min}|$ and initial guess $\mu_0 = 2 |\lambda_\text{min}|$. Each evaluation of $g$ requires solving the $3 \times 3$ linear system \eqref{eq:rational_3_linear_system}, and once the solution $\mu$ is found, we solve \eqref{eq:rational_3_linear_system} one more time to get the final values of $c_0$, $c_1$, and $c_2$.
%
%Row 1 corresponds to the condition $f(M)=M$, row 2 corresponds to the condition $f(0)=0$, and row 3 corresponds to the condition $f'(0)=0$. To find the third row of $A$, we used the fact that the derivative of $f$ is given by $$f'(\lambda) = c_1 - \frac{c_2}{(\lambda+\mu)^2}.$$
%
%
%We want $f(\lambda)$ to have the following desirable properties:
%\begin{enumerate}
%	\item $f(\lambda)$ is nonnegative on $[m, M]$.
%	\item $f(\lambda) \approx \lambda$ when $\lambda$ is positive and not small.
%	\item $f(0)=0$ is the minimum of $f(\lambda)$ on $[m, M]$.
%	\item $f(\lambda) \le |\lambda|$ when $\lambda$ is negative.
%\end{enumerate}
%Property 1 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. Property 2 ensures that $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. That is, the part of the spectrum of $\mathcal{A}$ that we expect to be well-approximated by the product-convolution approximation. Property 3 ensures that $\widetilde{\mathbf{A}}_\text{sym}^+$ preserves the null-space (and near null-space) of $\widetilde{\mathbf{A}}_\text{sym}$. This is important because in practice, we often want to approximate operators $\mathcal{A}$ that have a large (possibly infinite) number of zero eigenvalues, and/or a sequence of eigenvalues that clusters at zero. Hessians in ill-posed inverse problems typically have both of these properties. Property 4 ensures that $f$ does not amplify erroneous negative eigenvalues $\widetilde{\mathbf{A}}_\text{sym}$. If $\mathcal{A}$ is positive semi-definite, then eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ can only be negative because of errors in the product-convolution approximation.
